Circular Distributional Trees and Forests with an Application to Wind Direction Forecasting

### 2019-11-21 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of Table 1; resolve FIXMEs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2890 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 1117b24c4e4d8609763cdeab7469e40c9ebf3b80
Update draft (Lisa and Moritz 'Proofread')
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2892 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8e73c665f95f6ef08f1cc7e70cc0d9cf65a86ea6

\Abstract{
As circular data appears in many different areas, the development of appropriate statistical 
models which also account for periodicity has become an important but challenging research field. 
For probabilistic modeling of circular data typically the von Mises distribution is employed.
In order to regress its parameters on covariates, a circular distributional tree is proposed as 
an alternative to more commonly-used additive models. The resulting circular trees are 
easy to interpret, can detect non-additive effects, and select covariates and their 
interactions automatically.
By combining an ensemble of circular trees a more stable and regularized circular distributional forest
can be built which allows for modeling smooth effects as well.
%To stabilize and regularize the model an ensemble of distributional trees can be combined 
%yielding a distributional forest which allows for modeling smooth effects as well.

@@

%In order to assess the predictive performance of the proposed circular trees and forests,
%a case study on probabilistic forecasting of wind direction for two airports in Austria is presented
%with alternative commonly used approaches being employed as benchmarks.

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
the von Mises distribtuion is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. 
%There, within an iterative 
%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
%To obtain proper initial values it is recommended to apply two separate models where either one of 
%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology 
%and the contained optimization problem have 
has shown to be very challenging in their application, mainly due to the 
complexity encountered in optimizing the log-likelihood function which is not globally concave 
\citep{Pewsey+Neuhaeuser+Ruxton:2013, Gill+Hangartner:2010}. Therefore, 
highly informative starting values are crucial for the model to converge at all.
In order to avoid this strong dependence on appropriate initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach to regression modeling of circular data by employing a GLM 
in a Bayesian framework. 
%\fixme{revise literature survey with focus on transition to GLM}
The named methods and also many of their further developments \citep[see,
e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive regression models.
While generally providing a powerful toolbox for modeling particularly smooth
effects, they may not be able to capture all types of effects adequately.
Especially abrupt changes might not be detected properly owing to the enforced
additive structure and model assumptions such as a possibly highly non-linear link function and
the form of linear combinations.

An alternative modeling approach that can deal with non-additive effects is provided
by tree-based methods. \cite{Lund:2002} introduced the concept of circular regression trees 
which was further extended by \cite{Larraondo+Inza+Lozano:2018}. 
Following this idea of employing tree-based methods but also including the probabilistic concept
of regressing distribution parameters on covariates, we propose a very flexible approach 
by applying the von Mises distribution within the methodology of distributional trees and 
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular data.
In that way no model assumptions are required as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while combining an 
ensemble of trees to a forest model allows for modeling smooth changes as well. 
Thus, a high level of flexibility is provided while a full distributional model is specified 
by distributional trees and forests such that a wide range of inference is made available.

@@

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for wind direction forecasting 
at the airport of Innsbruck for the following hour.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

A more extensive case study of circular distributional trees and forests applied to
probabilistic forecasting of wind direction at the airports of Innsbruck and
Vienna will be presented in this paper. Moreover, their performances are
benchmarked to alternative commonly used approaches.  The theory on
probabilistic circular modeling introducing the von Mises distribution will be
discussed in Section~\ref{sec:prob_circ}. The methodology of circular
distributional trees and forests and their features are introduced in
Section~\ref{sec:tree_forest}.  After the case study presented in
\ref{sec:wind}, a comprehensive summary and conclusion is given in
Section~\ref{sec:summary}.

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

Once an appropriate distribution family is selected fitting a probabilistic model corresponds to
estimating the distribution parameters. Thus, when applying the von Mises distribution to fit a 
probabilistic model $\text{vM}(Y, \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
the distribution parameters $\mu$ and $\kappa$ are to be estimated.
For a learning sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by 
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim}
(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i),
\end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that a fully specified 
distributional model is fit to the learning data.

By means of the score function
\begin{equation}
\label{eq:scores}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
for each observation $y_i$. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
%\centering
%\hspace{-1cm}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \vspace{-1em}
  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
\end{minipage}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\caption{Illustration of a distributional model fitted to circular data by 
  maximum likelihood-estimation employing the von Mises distribution. 
  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
  Right panel: circular illustration.
  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
  with the estimated location parameter (red hand).
  }
  \label{fig:densities} 
  %\includegraphics[width=4cm]{density_vM.pdf}
\end{figure}
%\fixme{update figure 2d to 3d, using data of example tree}

Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
employing the von Mises distribution. While the left panel only serves as an illustration 
of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
consequently not accounting for periodicity the right panel displays an appropriate 
representation of a circular quantity on the unit circle. 
% on observed wind direction data from the airport of Innsbruck.
Looking at the fitted density (red line) and estimated location (red hand) this example 
shows that reasonably well fitting probabilistic models can be established for circular 
data by specifying an appropriate distribution family such as the von Mises distribution
and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable $Y$. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Circular distributional trees and forests}
\label{sec:tree_forest}
%\fixme{revise transition from sec\,2. to sect\,3. with the focus on circular methods}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and specifying a model 
properly can be a very challenging, particularly for a high number of covariates and no information
on possible interactions.
An alternative regression approach that can deal with these problems is to apply a tree-structured 
model.

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a better 
fitting overall model.
This is the general idea of regression trees which are combined with distributional regression
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
By applying the presented distributional trees a full distributional model is specified in each 
node of the tree. For this purpose a distribution family has to be chosen in advance. Hence, 
selecting the von Mises distribution allows for an application of distributional trees to circular
data following the tree building algorithm as introduced in Section~2.2. of 
\cite{Schlosser+Hothorn+Stauffer:2019}.

In particular, in each node the crucial decision on how and where to split the data is based on model
scores which are obtained by evaluating the score function $s$ as defined in Equation~\ref{eq:scores} 
at the individual observations and parameter estimates. For the von Mises distribution and a data set 
of $n$ observations this yields an $n \times 2$ matrix that can be employed as a kind of residual,
capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
$\hat{\kappa}$, respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

Once a distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

@@

\subsection{Circular forests}
\label{sec:circforest}
A natural extension of distributional trees for circular data are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Hence, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
splits from the tree.
Thus distributional forests for circular data can capture both, smooth and abrupt changes, while
covariates and possible interactions are selected automatically allowing for an easy to
apply methodology to fully specify the von Mises distribution for circular data.

>>>>

\Abstract{
As circular data appears in many different areas, the development of appropriate statistical 
models which can account for periodicity has become an important but challenging research field. 
For probabilistic modeling of circular data typically the von Mises distribution is employed.
In order to regress its parameters on covariates, a circular distributional tree is proposed as 
an alternative to more commonly-used additive models. The resulting circular trees are 
easy to interpret, can detect non-additive effects, and select covariates and their 
interactions automatically.
By combining an ensemble of circular trees a more stable and regularized circular distributional forest
can be built which allows for modeling smooth effects as well.

@@

@@

In many approaches to model circular data it is assumed that the circular variable of interest 
follows a circular distribution, in particular the von Mises distribution which is also known 
as the ``circular normal distribution'' (see Section~\ref{sec:prob_circ} for more details). 
This is also the basis which early regression models for circular data are built on as they 
regress the parameter(s) of the von Mises distribution on covariates.
One of the first regression models with a circular response variable and linear covariates
was presented by \cite{Gould:1969} where the circular mean, i.e., the location parameter of 
the von Mises distribtuion is predicted by a linear combination of covariates.
\cite{Johnson+Wehrly:1978} took this idea one step further by plugging in a link function such that the values of the 
linear predictor are transformed to a restricted interval of length $2\pi$. 
This generalized linear model (GLM)-type approach was further extended by \cite{Fisher+Lee:1992} and subsequently by \cite{Fisher:1993} 
introducing another specific transformation function which maps the values of the real line to the 
unit circle. Moreover, the presented methodology allows for both distribution parameters, the mean direction and the concentration parameter of the von Mises distribution, to be modeled depending
on covariates, not only in separate models but also in a mixed model. 
%There, within an iterative 
%optimization algorithm the two parameters are estimated and improved stepwise until convergence. 
%To obtain proper initial values it is recommended to apply two separate models where either one of 
%the parameters is fixed while the other is predicted employing covariates in a GLM-type model. 
While being built on well elaborated theory this methodology 
%and the contained optimization problem have 
has shown to be very challenging in their application, mainly due to the
complexity encountered in optimizing the log-likelihood function which is not
globally concave.  Therefore, highly informative starting values are crucial
for the model to converge at all \citep{Pewsey+Neuhaeuser+Ruxton:2013,
Gill+Hangartner:2010}.  In order to avoid this strong dependence on appropriate
initial values, \cite{Mulder+Klugkist:2017} presented an alternative approach
to regression modeling of circular data by employing a GLM in a Bayesian
framework.  The listed methods and most of their further developments
\citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} are built on additive
regression models.  While these generally provide a powerful toolbox for modeling
particularly smooth effects, they may not be able to capture all types of
effects adequately.  Especially abrupt changes might not be detected properly
owing to the enforced additive structure and model assumptions, such as a
possibly highly non-linear link function and the form of linear combinations.

An alternative modeling approach that can deal with non-additive effects is
provided by tree-based methods. \cite{Lund:2002} introduced the concept of
circular regression trees which was further extended by
\cite{Larraondo+Inza+Lozano:2018}.  Following this idea of employing tree-based
methods but also including the probabilistic concept of regressing distribution
parameters on covariates, we propose a very flexible approach by applying the
von Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019} to circular
data.  The resulting circular distributional trees and forests do not require
any model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure
while circular forests allow for modeling smooth changes as well.  Thus, a high
level of flexibility is provided while a full distributional model is specified offering
a wide range of inference.

@@

\begin{figure}[t]
\centering
\includegraphics[width = \textwidth]{_plot_circforest_finalexampletree_ibk_lag1_v14.pdf}
\caption{Fitted tree based on the von Mises distribution for 1-hourly wind direction forecasts 
at the airport of Innsbruck.
In each terminal node the empirical histogram (gray) and fitted density (red line)
are depicted along with the estimated location parameter (red hand). The covariates
employed are wind direction (degree), wind speed ($\text{ms}^{-1}$),
and pressure gradients ($\text{dpressure; hPa}$) west, east and south of the airport,
all lagged by one hour.}
\label{fig:tree_ibk} 
\end{figure}

A more extensive case study of circular distributional trees and forests
applied to probabilistic forecasting of wind direction at the airports of
Innsbruck and Vienna will be presented in this paper. Moreover, their
performances are benchmarked to alternative commonly used approaches.  The
remainder of the paper is structured as follows: The theory on probabilistic
circular modeling introducing the von Mises distribution will be discussed in
Section~\ref{sec:prob_circ}. The methodology of circular distributional trees
and forests and their features are introduced in Section~\ref{sec:tree_forest}.
After the case study presented in \ref{sec:wind}, a comprehensive summary and
conclusion is given in Section~\ref{sec:summary}.

\section{Probabilistic circular modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in (-\pi, \pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in (-\pi, \pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

Once an appropriate distribution family is selected fitting a probabilistic model corresponds to
estimating the distribution parameters. Thus, when applying the von Mises distribution to fit a 
probabilistic model $\text{vM}(Y; \mu, \kappa)$ to a circular response variable $Y \in \mathcal{Y}$ 
the distribution parameters $\mu$ and $\kappa$ are to be estimated.
For a learning sample with $n$ observations $\{y_i\}_{i=1,\ldots,n}$ this can be done by 
maximizing the log-likelihood function $\ell(\mu, \kappa; y)$, hence
\begin{equation}\label{eq:optim}
(\hat{\mu},\hat{\kappa}) = \argmax_{\mu,\kappa} \sum_{i=1}^n \ell(\mu, \kappa; y_i)
\end{equation}
yielding maximum likelihood estimators $\hat{\mu}$ and $\hat{\kappa}$ such that a fully specified 
distributional model is fit to the learning data.

By means of the score function
\begin{equation}
\label{eq:scores}
\begin{aligned}
s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
\smallskip &,
\quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y_i) \bigg) 
\\
&= \bigg(\kappa \sin(y-\mu)
\smallskip &,
\quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
\end{aligned}
\end{equation}
an indicator of deviation and in that way a measure of goodness of fit of the model can be obtained 
for each observation $y_i$ and fitted parameter. Moreover, the optimization problem in Equation~\ref{eq:optim} can also 
be defined by 
%setting the sum of the first derivative of the log-likelihood function evaluated for all observations to zero, i.e., by setting
\begin{equation}
\sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}

\begin{figure}[t]
%\centering
%\hspace{-1cm}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \vspace{-1em}
  \includegraphics[width=1.05\linewidth]{density_linear.pdf}
  \end{minipage}%
\begin{minipage}{.45\textwidth}
  %\centering
  \includegraphics[width=1.2\linewidth]{density_circular.pdf}
\end{minipage}
\begin{minipage}{.05\textwidth}
\hfill
\end{minipage}%
\caption{Illustration of a distributional model fitted to circular data by 
  maximum likelihood estimation employing the von Mises distribution. 
  Left panel: ``linear'' illustration restricted to the interval $[0, 2\pi)$. 
  Right panel: circular illustration.
  In both panels the empirical histogram (gray) and fitted density (red line) are depicted along 
  with the estimated location parameter (red hand).
  }
  \label{fig:densities} 
  %\includegraphics[width=4cm]{density_vM.pdf}
\end{figure}

Figure~\ref{fig:densities} depicts a distributional model fitted to circular data
employing the von Mises distribution. While the left panel only serves as an illustration 
of the corresponding  ``linear'' description restricted to an interval of length $2\pi$ but 
consequently not accounting for periodicity, the right panel displays an appropriate 
representation of a circular quantity on the unit circle. 
% on observed wind direction data from the airport of Innsbruck.
Looking at the fitted density (red line) and estimated location parameter (red hand) this example 
shows that reasonably well fitting probabilistic models can be established for circular 
data by specifying an appropriate distribution family such as the von Mises distribution
and estimating its parameters via maximum likelihood. 
However, this procedure considers only the response variable. Including covariates 
can improve any model remarkably as they might provide valuable additional information.
Therefore, distributional regression models are of high interest for all types of data,
including circular data.

\section{Circular distributional trees and forests}
\label{sec:tree_forest}
A common way of including covariates is to regress the model parameters on covariates.
In the case of a probabilistic model for a circular response this corresponds to expressing the 
location parameter $\mu$ and the concentration parameter $\kappa$ of the von Mises distribution
by means of the available covariates, e.g., in the form of a linear combination, possibly with an 
appropriate link function plugged in as in GLMs. This kind of approach is presented for example
in the mixed model of \cite{Fisher+Lee:1992} and \cite{Fisher:1993}. 
%While applying the mixed model 
%requires expert user knowledge due to its complexity as discussed in Section~\ref{sec:introduction},
%\cite{Mulder+Klugkist:2017} tried to facilitate applications by their Bayesian regression model for
%a circular response variable.
Apart from the methodology developed specifically for circular data more general approaches to 
regressing model parameters on covariates are commonly used such as for example generalized additive 
models for location, scale, and shape \citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005} 
where each distribution parameter is estimated by its own generalized additive model (GAM). 
%However, an adaption of this methodology to circular data can be very complex due to a
%challenging representation of smooth effects on the unit circle.
%However, an adaption of this methodology to circular data can not be obtained by the straight 
%forward approach of simply wrapping the modeled additive effects
%around the unit circle since the parameters of the von Mises distribution and their 
%dependence/interaction differs strongly from the linear case and therefore requires specific handling.
However, difficulties might arise in case of non-additive changes and a proper model 
specification can be very challenging, particularly for a high number of covariates and no information
on possible interactions.
As an alternative regression approach which can deal with these problems, we introduce a circular probabilistic tree-structured model in this section.

\subsection{Circular trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly for 
complex data with strong variations. Therefore, separating the data set into more homogeneous 
subgroups based on covariates before fitting a local model in each of these subgroups allows 
to capture (possibly) group specific effects more precisely and hence can result in a overall better 
fitting model.
This is the general idea of regression trees which are combined with distributional 
modeling in \cite{Schlosser+Hothorn+Stauffer:2019} based on the unbiased recursive partitioning 
algorithms MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree \citep{Hothorn+Hornik+Zeileis:2006}.
By applying the presented distributional trees, a full distributional model is specified in each 
node of the tree for a specific distribution family chosen in advance. Hence, 
selecting the von Mises distribution allows for an application of distributional trees to circular
data following the tree building algorithm as introduced in Section~2.2. of 
\cite{Schlosser+Hothorn+Stauffer:2019}.

In the applied tree-building algorithm, in each node the crucial decision on how and where to split the data is based on model
scores which are obtained by evaluating the score function $s$ 
for the individual observations and parameter estimates (Equation~\ref{eq:scores}). For the von Mises distribution with its two distribution parameters and a data set 
of $n$ observations this yields an $n \times 2$ matrix that can be employed as a discrepancy measure,
capturing how well each given observation conforms with the estimated location $\hat{\mu}$ and precision
$\hat{\kappa}$, respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test~(MOB) or a permutation test~(CTree).
By doing so in each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

Once a distributional tree model is fitted it can be employed to obtain probabilistic predictions
for a possibly new set of observed covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

@@

\subsection{Circular forests}
\label{sec:circforest}
A natural extension of circular distributional trees are ensembles or forests 
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by \cite{Breiman:2001} average the predictions of an ensemble
of trees, each built on a subsample or bootstrap of the original data. 
A generalization of this strategy is to obtain weighted predictions by adaptive local 
likelihood estimation of the distributional parameters 
\citep[Section~2.3. of][]{Schlosser+Hothorn+Stauffer:2019, Hothorn+Zeileis:2017}. 
More specifically, for each possibly new observation~$\bm{z}$ a set of averaged ``nearest neighbor''
weights $w_i^{forest}(\bm{z})$ 
%\citep{Lin+Jeon:2006}
is obtained that is based on the number of trees in which $\bm{z}$ is assigned to the same terminal 
node as each learning observation $y_i, i \in \{1,\ldots,n\}$.
Hence, for a forest of $T$ trees the weights are calculated via
\begin{equation}
w^{\text{forest}}_i(\bm{z}) = \frac{1}{T} \sum_{t=1}^T \sum_{b=1}^{B^t}
\frac{\mathbf{1}((\bm{z}_i \in \mathcal{B}^t_b) \land (\bm{z} \in \mathcal{B}^t_b))}{|\mathcal{B}^t_b|},
\end{equation}
where $|\mathcal{B}^t_b|$ denotes the number of observations in the $b$-th
segment of the $t$-th tree.
Therefore, similar observations ending up more often in the same terminal node have higher weights
and in that way a stronger influence in the estimation process.

@@

Therefore, the resulting parameter estimates can smoothly adapt to the given
covariates $\bm{z}$ whereas $w_i^{forest}(\bm{z}) = 1$ would correspond to the unweighted
full-sample estimates and $w_i^{forest}(\bm{z}) \in \{0, 1\}$ corresponds to the abrupt
splits from a tree.
Thus, circular distributional forests can capture both smooth and abrupt changes, while
covariates and possible interactions are selected automatically allowing for an easy-to-apply 
methodology with a full specification of the von Mises distribution.

\section{Case study: Probabilistic wind direction forecasting}
\label{sec:wind}
Accurate forecasts of wind directions are of great importance for risk
management in various fields such as agriculture, energy production or air safety control. 
For example, in order to direct airplanes to a safe landing, precise knowledge of wind 
direction for the next hour(s) at the respective airport is highly desirable and adequate 
prediction methods are required. In this section, a comprehensive case study on probabilistic
wind direction forecasting is provided for two airports in Austria with fundamentally
different site characteristics: Innsbruck with a rather unusual location for an
airport, located at the bottom of a deep Alpine valley, and Vienna located
within flat terrain in the East of Austria. In addition to the newly
introduced circular distributional trees and forests, the case study employs various
alternative approaches to forecast wind direction as a benchmark. The study is based on
$1$\,h and $3$\,h forecasts employing lagged observations in the vicinity of
the airports as predictor variables.
